---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Introduction
This document contains the **Reproducible Research** assignment for Week #2 at **John Hopkins University**'s DataScience course.

## Loading and preprocessing the data
The DataSet is provided in a ZIP file. The first task consists in decompress the file, load the DataSet and prepare the data.

```{r load-data, cache=TRUE}
# Unzip the file
unzip( "activity.zip" )

# DataSet load
activity_data <- read.csv( "activity.csv" )
```


The `interval` column must be interpreted as a *factor* instead of an integer.

```{r transcform-data, cache=TRUE}
# Convert 'interval' to factor
activity_data[,'interval'] <- factor( activity_data[,'interval'] )
```


We will keep the original DataSet in the `activity_data` DataFrame and a clean set (without 'NA' values) in the `complete_data` DataFrame.

```{r clean-data, cache=TRUE}
# Remove the NA entries
complete_dataset <- activity_data[ complete.cases( activity_data ), ]
```


## What is mean total number of steps taken per day?

Let's plot a histogram about the daily steps.

```{r steps-by-day-histogram, cache=TRUE}
# Aggregate the steps (by day)
steps_by_day <- aggregate( steps ~ date, complete_dataset, FUN=sum )

# Plot a histogram (steps by day) using the complete dataset
hist(steps_by_day$steps,
     main="Steps by day (histogram)",
     xlab="Steps",
     breaks=10,
     col="grey")
```

Now let's get the Mean and Median.

```{r mean-median, cache=TRUE}
# Compute the data summary
steps_by_day_summary <- summary( steps_by_day$steps )

# The mean
steps_by_day_mean <- steps_by_day_summary["Mean"]

# The median
steps_by_day_median <- steps_by_day_summary["Median"]
```

The **Mean** is **`r toString(as.integer(steps_by_day_mean))`** and the **Median** **`r toString(steps_by_day_median)`**.

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
